---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the data

The Redfield Ratio was defined by observations made by scientist Albert Redfield that across all ocean environments, the ratio of carbon, nitrogen, and phosphorus in the oceans (C/N/P) was the same - (106:16:1). There is also a latitudinal effect where, in lower latitudes that are warm and nutrient-depleted, a ratio of 195:28:1 is observed, while in cold, nutrient-rich high latitude regions the ratio observed is 78:13:1. The Redfield Ratio can therefore be used to determine how balanced an ocean ecosystem is. We can then use this to ask questions about how aquatic life may have evolved in response to a changing global temperatures and nutrient availability.

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course). This seawater chemistry dataset specifically looks at data from the Arctic Ocean.

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read.csv(here::here("data","raw", "BGchem2008data.csv")) # use ../ to go up one directory
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
# str(bg_chem)
 head(bg_chem)
# summary(bg_chem)
```

# Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

## Calculate summary statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
```

## Calculate mean Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
ratio
```

You can report results inline in Quarto like this:

The Redfield ratio for this dataset is approximately: `r round(ratio)`

## Plot Redfield ratio measurements

Finally, create a simple base R plot that shows the relationship between phosphate and the total nitrogen components.

```{r}
plot(
  bg_chem$P,
  bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4,
  xlab = "Phosphate (P)",
  ylab = "Total nitrogen (NO2 + NO3 + NH4)",
  main = "Ocean water samples: total N vs P"
)
```

This plot depicts the ratio of nitrogen to phosphate, and demonstrates an almost linear correlation between the two such that as nitrogen levels increase, so do phosphate levels. It also shows how when there is low nitrogen, there is also low phosphorus in the Arctic Ocean seawater. Interestingly, there are many sites where low nitrogen and phosphorus were observed. There are also a few sites where the standard Redfield Ratio of 16:1 (N:P) is not observed in the Arctic Ocean, with ratios closer to 12:1 or even 3:1, a little more consistent with the Redfield Ratio observed in cold, high latitude regions. However, there are also many sites where the Redfield Ratio is close to expected, about 16:2 N:P.

# Conclusion

In this mini-analysis, we:

-   Loaded packages in a dedicated setup chunk
-   Read in `BGchem2008data.csv` using a project-root-relative path (`here()`)
-   Calculated simple nutrient means
-   Computed a mean nitrogen:phosphate ratio
-   Created a basic plot relating total nitrogen to phosphate

This is the core idea of **literate analysis**: combine narrative and code in one document so the workflow is easier to understand and reproduce.
